---
title: "Cleaning Data"
format: html
---

## Setup

Setting up the libraries.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(readr)
```

## Import

Importing the raw beverage data.

```{r}
bev_raw <- read_csv("data-original/bev-raw.csv")

bev_raw |> head()
```

## Cleaning columns

Selecting the columns relevant to analysis.

```{r}
bev_col <- bev_raw |>
  select(
    `Taxpayer Number`,
    `Location Number`,
    `Location Name`,
    `Location Address`,
    `Location City`,
    `Location Zip`,
    `Obligation End Date`,
    `Beer Receipts`,
    `Wine Receipts`,
    `Liquor Receipts`,
    `Total Receipts`
  ) |>
  clean_names()

bev_col
```

## Convert data types

Fixing the data types.

```{r}
bev_clean <- bev_col |>
  mutate(
    date = mdy(obligation_end_date),
    year = year(date),
    location_zip = as.character(location_zip),
    beer_receipts = as.numeric(gsub("[,$]", "", beer_receipts)),
    wine_receipts = as.numeric(gsub("[,$]", "", wine_receipts)),
    liquor_receipts = as.numeric(gsub("[,$]", "", liquor_receipts)),
    total_receipts = as.numeric(gsub("[,$]", "", total_receipts))
  ) |>
  filter(year != 2026)

bev_clean
```

## Export

Export the cleaned dataset.

```{r}
write_rds(bev_clean, "data-processed/bev_clean.rds")
```

The use of gsub() in this notebook was informed by [this website](https://www.r-bloggers.com/2024/03/mastering-text-manipulation-in-r-a-guide-to-using-gsub-for-multiple-pattern-replacement/) after my code failed to run and I tried numerous methods to change my code. The original problem was different but I ended up forgetting to switch this back to parse_number(). 

