---
title: "Data Analysis"
format: html
---

## Goal of this notebook

This notebook analyzes alcohol sales trends in Austin after COVID-19, specifically looking at recovery patterns across locations and drink types.

## Setup

Setting up the libraries.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(lubridate)
library(readr)
```

## Import

Importing the data.

```{r}
#| label: import

bev_clean <- read_rds("data-processed/bev_clean.rds")

bev_clean |> glimpse()
```

## Overall sales trend over time

Exploring how total alcohol sales in Austin changed over time, specifically before and after COVID-19.

```{r}
annual_sales <- bev_clean |>
  filter(year != 2026) |>
  group_by(year) |>
  summarize(
    total_sales = sum(total_receipts, na.rm = TRUE)
  ) |>
  arrange(year)

annual_sales
```

Data Takeaway: Alcohol sales in Austin dropped sharply during the COVID-19 pandemic, falling from about $883 million in 2019 to $426 million in 2020, a decline of more than half. Sales rebounded in 2021 to roughly $859 million, nearly returning to pre-pandemic levels. In the years that followed, total sales continued to grow, surpassing pre-COVID figures and reaching about $1.14 billion in 2025, the highest in the dataset.

## Monthly trend around COVID

Comparing monthly revenue trends before, during and after the pandemic.

```{r}
monthly_sales <- bev_clean |>
  filter(year != 2026) |>
  group_by(year, month = month(date)) |>
  summarize(
    total_sales = sum(total_receipts, na.rm = TRUE)
  ) |>
  arrange(year, month)

monthly_sales
```

Data Takeaway: Monthly alcohol sales in Austin followed a stable  pattern in 2019, ranging from about $60 million to $89 million per month. In 2020, sales dropped sharply beginning in March, falling to just around $2.7 million in April, the lowest point in the dataset. Sales gradually recovered throughout the year and returned to more typical levels by late 2020. By 2021 and 2022, monthly sales had fully rebounded, with the highest months reaching over $100 million, and remained strong through 2025.

## Zip code differences

Recovery comparison by ZIP code, using 2019 as a baseline before, 2020 as sales drop baseline and 2025 as the recovery baseline.

```{r}
zip_summary <- bev_clean |>
  filter(year != 2026) |>
  group_by(location_zip, year) |>
  summarize(total_sales = sum(total_receipts, na.rm = TRUE)) |>
  group_by(location_zip) |>
  summarize(
    sales_2019 = total_sales[year == 2019],
    sales_2020 = total_sales[year == 2020],
    sales_2025 = total_sales[year == 2025],
    recovery_2025 = sales_2025 / sales_2019,
    covid_drop = sales_2020 / sales_2019
  ) |>
  arrange(desc(recovery_2025))

zip_summary
```

Data Takeaway: Alcohol sales recovery after COVID-19 varied across Austin-area ZIP codes. Some areas more than fully rebounded by 2025, with ZIP code 78733 reaching more than 18 times its 2019 sales level, while others remained below pre-pandemic levels. Overall, most ZIP codes exceeded their 2019 baseline, but recovery was uneven across the city.

Downtown and core Austin ZIP codes such as 78701 recovered more slowly, reaching about 1.1 times their 2019 sales level by 2025 after a sharp decline in 2020. In contrast, several suburban ZIP codes saw much stronger growth, including 78733 and 78754, which experienced some of the highest recovery rates in the dataset. A small number of ZIP codes, including 78757 and 78750, remained slightly below their pre-COVID levels by 2025, suggesting slower long-term recovery in parts of North and Northwest Austin.

note to self: ask prof if this is too long? feels like i got good data from here but #idk

## Establishment recovery

Seeing which establishments recovered the best.

```{r}
top_locations <- bev_clean |>
  filter(year != 2026) |>
  group_by(location_name, location_address, year) |>
  summarize(
    total_sales = sum(total_receipts, na.rm = TRUE),
  )

location_summary <- top_locations |>
  group_by(location_name, location_address) |>
  summarize(
    sales_2019 = total_sales[year == 2019],
    sales_2020 = total_sales[year == 2020],
    sales_2025 = total_sales[year == 2025],
    .groups = "drop"
  ) |>
  filter(
    !is.na(sales_2019),
    sales_2019 > 0,
    !is.na(sales_2025)
  ) |>
  mutate(
    recovery_2025 = sales_2025 / sales_2019,
    covid_drop = sales_2020 / sales_2019
  )

location_summary |>
  arrange(desc(recovery_2025)) |>
  head(10)
```

Data Takeaway: Austin’s highest-grossing alcohol establishments showed different recovery patterns after the COVID-19 pandemic. Several places rebounded strongly by 2025 and even exceeded their pre-pandemic levels. Proper Hotel rose from about $104,000 in 2019 to $10.8 million in 2025, while Scoot Inn Concession Company increased from roughly $807,000 to $14.9 million over the same period.

Other venues saw more moderate but still strong growth, including Mala Vida, which grew from about $69,000 to $3.36 million, and Aloft Austin Airport, which increased from about $2,300 to $225,000. In contrast, established venues such as Fairmont Austin Hotel and Buford’s recovered more slowly and remained closer to or below their 2019 levels by 2025.

## Alcohol types over time

Comparison of beer, wine, and liquor sales using 2019 as a baseline before, 2020 as sales drop baseline and 2025 as the recovery baseline.

```{r}
drink_trends <- bev_clean |>
  filter(year != 2026) |>
  group_by(year) |>
  summarize(
    beer = sum(beer_receipts, na.rm = TRUE),
    wine = sum(wine_receipts, na.rm = TRUE),
    liquor = sum(liquor_receipts, na.rm = TRUE)
  )

drink_trends
```

Data Takeaway: All three alcohols dropped sharply in 2020, but recovered by 2021 and continued to grow through 2025. Beer and liquor showed the strongest post-COVID growth, while wine remained the smallest and most stable category over time.

note to self: is this worth it to include? no strong conclusions that i can tie back to my original question

